---
title: "Assignment - 1"
author: "Soni Verma"
date: "2024-01-31"
output: html_document
---
### 1
```{r, echo = F}
# Change the values of N and theta as needed
N <- 50
theta <- 0.4

# Define the CDF of the binomial distribution
cdf <- function(x) {
  pbinom(N * x, size = N, prob = theta)
}

# Define the CDF derived from the Central Limit Theorem
ass_cdf <- function(y) {
  pnorm(y, mean = N * theta, sd = sqrt(N * theta * (1 - theta)))
}

# Plot the CDF and CLT-CDF
plot(cdf, from = 0, to = 1, col = "red", lty = 1, ylab = "Probability")
plot(ass_cdf, from = 0, to = 1, add = TRUE, col = "purple", lty = 3)
legend("topleft", legend = c("CDF", "CLT-CDF"), col = c("red", "purple"), lty = c(1, 3))

# Evaluate the maximum absolute difference between the two CDFs
support <- seq(0, 1, by = 1/N)
result <- max(abs(cdf(support) - ass_cdf(support)))

# Print the result
cat("Maximum absolute difference between CDF and CLT-CDF:", result, "\n")

```


### Question 2
Distribution of $X_n$ is coming out to a Gamma function with shape parameter $N\alpha$ and rate parameter $N\lambda$.

### Question 3

To check f(x) to be a valid pdf , we have to check two condition: 

- It should be non negative for its entire range.

- On integrating it over its entire range should equal 1e. 

```{r, echo = FALSE, include = FALSE}
library(pracma)

# Defining integrand function
integrand <- function(x) {
  f <- 2*besselK( 2*sqrt(x), nu = 0)  
  return (f)
}

# Perform numerical integration
result <- integrate(integrand, lower = 0, upper = Inf)
```

 
```{r, echo = TRUE}
print(result)
```

As we can see, integral value of bessel function on integrating over given space is coming out to be 1 which implies f(x) is a valid pdf. 

### Question 4
- Joint distribution of X is coming out to be product of  marginal distribution of $X_1$ and product of all conditional distribution from $X_2$ to $X_T$.
- Conditional distribution of $X_t$ given $X_1$...$X_T$ is only depending on just previous point $X_(t-1)$ . Therefor, it would be normally distributed.


### Question 6

By applying Bayes theorem, we get that posterior distribution is coming out to be gamma distribution.


### Question 7

Firstly writing function for given curve ,then applying accept reject for generating samples.

```{r , include = FALSE, echo = F}
library(tictoc)
```

```{r setup, include=FALSE}
# Function to check if a point is inside the region
internal_points <- function(x, y) {
  (x^2 + y^2 - 1)^3 <= x^2 * y^3
}

# Function to generate 10000 IID samples using acceptance-rejection sampling
generate_samples <- function(n) {
  samples <- matrix(NA, ncol = 2, nrow = n)
  count <- 0
  
  while (count < n) {
    x <- runif(1, -2, 2)
    y <- runif(1, -2, 2)
    
    if (internal_points(x, y)) {
      count <- count + 1
      samples[count, ] <- c(x, y)
    }
  }
  
  return(samples)
}
```

- Cheking time colapsed in generate_samples function using tictoc function using library tictoc.

```{r, echo = FALSE}
set.seed(100)

# Start the timer
tic()

# Generating IID samples
samples <- generate_samples(1e4)

# Stop the timer
elapsed_time <- toc()
```

```{r, echo = FALSE}
# Plotting samples and the region
plot(samples, col = "purple", pch = 16, main = "Acceptance-Rejection Sampling")

```
